Certified Symbolic Finite Transducers: Formalization and Applications to String Analysis

### 2025-09-02 Paper/application.tex
polish application

An important application of our formalization lies in string sanitization for web applications. Sanitizers are crucial for preventing security vulnerabilities such as cross-site scripting (XSS) attacks by ensuring that user inputs are properly encoded before rendering in web browsers. Listing \ref{lst-html-encode} presents a C\# code snippet implementing an HTML encoding function—a fundamental sanitization operation. This function iterates through each character in the input string and encodes characters that are unsafe for HTML rendering.

\begin{lstlisting}[language={[Sharp]C}, caption={C\# Code for AntiXSS.EncodeHtml version 2.0.}, label={lst-html-encode}, float=htbp]
static string EncodeHtml(string t)
{
  if (t == null) { return null; }
  if (t.Length == 0) {
    return string.Empty;
  }
  StringBuilder builder = new
    StringBuilder("", t.Length * 2);
  foreach (char c in t) {
    if ((((c > ''') && (c < '{')) |$\mid\mid$|
         ((c > '@') && (c < '['))) |$\mid\mid$|
         (((c == ' ') || ((c > '/') && (c < ':'))) |$\mid\mid$|
         (((c == '.') || (c == ',')) |$\mid\mid$|
         ((c == '-') || (c == '_'))))) {
          builder.Append(c);
      }
      else {
          builder.Append("&#" +
          ((int) c).ToString() + ";");
      }
  }
  return builder.ToString();
}
\end{lstlisting}

Figure \ref{fig:html-sanitizer-sft} illustrates the SFT representation of the HTML encoding sanitizer function. The transducer accepts any input string and produces an output where safe characters are preserved using the identity function $\texttt{id}$, while unsafe characters are encoded as HTML entities. The set of safe characters is defined as $\texttt{safe\_chars} = [\texttt{' '}, \texttt{','}, \texttt{'.'}, \texttt{'-'}, \texttt{'\_'}] \cup [\texttt{'0'}-\texttt{'9'}] \cup [\texttt{'A'}-\texttt{'Z'}] \cup [\texttt{'a'}-\texttt{'z'}]$, which can be efficiently represented using the interval algebra we formalized. Unsafe characters are transformed using the encoding function $\texttt{encode}(c) = \texttt{``\&\#''} + \texttt{toString}(\texttt{ord}(c)) + \texttt{``\;''}$. When the alphabet is set to Unicode $[\mathtt{0x0000}, \mathtt{0x10FFFF}]$, the complement $\neg\texttt{safe\_chars}$ represents all Unicode characters outside the safe set, which can also be efficiently represented using our formalized interval operations.
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[shorten >=1pt, node distance=2.5cm, on grid, auto]
  % States
  \node[state, initial, accepting] (q0) {$q_0$};
  \node[state] (q1) [right=3.5cm of q0] {$q_1$};
  \node[state] (q2) [below=of q0] {$q_2$};

  \path[->] 
    % Transition for safe characters
    (q0) edge[bend left=20] node[above,midway=1] {$\texttt{safe\_chars}, \texttt{id}$} (q1)
    % Transition for unsafe characters  
    (q0) edge[bend right=20] node[left,midway] {$\neg\texttt{safe\_chars}, \texttt{encode}$} (q2)
    % Epsilon transitions back to q0
    (q1) edge[bend left=30] node[below,midway] {$\varepsilon, \lambda\varepsilon$} (q0)
    (q2) edge[bend right=20] node[right,midway] {$\varepsilon, \lambda\varepsilon$} (q0);
\end{tikzpicture}
\caption{SFT representation of the HTML encoding sanitizer function.}
\label{fig:html-sanitizer-sft1}
\end{figure}

We evaluate the performance of our SFT product operation by computing the product of the HTML encoding sanitizer SFT with SFAs constructed from randomly generated Unicode user inputs. This experimental setup reflects realistic usage scenarios where diverse character sets and input patterns are encountered. The experimental evaluation was conducted on a laptop with an Apple M4 processor and 24 GB of memory, with a one-minute time limit per test.

@@

We also compare our approach with two state-of-the-art string solvers: Ostrich \cite{pacmpl/ChenFHHHKLRW22} and CVC5 \cite{cvc5}.
The experimental results demonstrate that string replacement operations, particularly replace-all operations, pose significant challenges for existing state-of-the-art solvers.
The "unknown" label in the table indicates instances where CVC5 cannot determine satisfiability. CertiStrR outperforms both Ostrich and CVC5 across most benchmark subgroups, with the notable exception of the "Replace Str" category where CVC5 achieves better performance (though it experiences one timeout). CertiStrR demonstrates particularly strong performance in the "Replace All" category.

@@

Notably, PCP and RNA benchmarks cannot be solved by most state-of-the-art string
solvers, including CVC4 \cite{cvc4} and Z3 \cite{Z3-str3}, yet CertiStrR solves
them efficiently. Among existing solvers, only OSTRICH \cite{pacmpl/ChenFHHHKLRW22} can handle these benchmarks.

>>>>

An important application of our formalization lies in string sanitization for web applications. Sanitizers are crucial for preventing security vulnerabilities such as cross-site scripting (XSS) attacks by ensuring that user inputs are properly encoded before rendering in web browsers.

% Listing \ref{lst-html-encode} presents a C\# code snippet implementing an HTML encoding function—a fundamental sanitization operation. This function iterates through each character in the input string and encodes characters that are unsafe for HTML rendering.

% \begin{lstlisting}[language={[Sharp]C}, caption={C\# Code for AntiXSS.EncodeHtml version 2.0.}, label={lst-html-encode}, float=htbp]
% static string EncodeHtml(string t)
% {
%   if (t == null) { return null; }
%   if (t.Length == 0) {
%     return string.Empty;
%   }
%   StringBuilder builder = new
%     StringBuilder("", t.Length * 2);
%   foreach (char c in t) {
%     if ((((c > ''') && (c < '{')) |$\mid\mid$|
%          ((c > '@') && (c < '['))) |$\mid\mid$|
%          (((c == ' ') || ((c > '/') && (c < ':'))) |$\mid\mid$|
%          (((c == '.') || (c == ',')) |$\mid\mid$|
%          ((c == '-') || (c == '_'))))) {
%           builder.Append(c);
%       }
%       else {
%           builder.Append("&#" +
%           ((int) c).ToString() + ";");
%       }
%   }
%   return builder.ToString();
% }
% \end{lstlisting}

% Figure \ref{fig:html-sanitizer-sft} illustrates the SFT representation of the HTML encoding sanitizer function. The transducer accepts any input string and produces an output where safe characters are preserved using the identity function $\texttt{id}$, while unsafe characters are encoded as HTML entities. The set of safe characters is defined as $\texttt{safe\_chars} = [\texttt{' '}, \texttt{','}, \texttt{'.'}, \texttt{'-'}, \texttt{'\_'}] \cup [\texttt{'0'}-\texttt{'9'}] \cup [\texttt{'A'}-\texttt{'Z'}] \cup [\texttt{'a'}-\texttt{'z'}]$, which can be efficiently represented using the interval algebra we formalized. Unsafe characters are transformed using the encoding function $\texttt{encode}(c) = \texttt{``\&\#''} + \texttt{toString}(\texttt{ord}(c)) + \texttt{``\;''}$. When the alphabet is set to Unicode $[\mathtt{0x0000}, \mathtt{0x10FFFF}]$, the complement $\neg\texttt{safe\_chars}$ represents all Unicode characters outside the safe set, which can also be efficiently represented using our formalized interval operations.
% \begin{figure}[htbp]
% \centering
% \begin{tikzpicture}[shorten >=1pt, node distance=2.5cm, on grid, auto]
%   % States
%   \node[state, initial, accepting] (q0) {$q_0$};
%   \node[state] (q1) [right=3.5cm of q0] {$q_1$};
%   \node[state] (q2) [below=of q0] {$q_2$};

%   \path[->] 
%     % Transition for safe characters
%     (q0) edge[bend left=20] node[above,midway=1] {$\texttt{safe\_chars}, \texttt{id}$} (q1)
%     % Transition for unsafe characters  
%     (q0) edge[bend right=20] node[left,midway] {$\neg\texttt{safe\_chars}, \texttt{encode}$} (q2)
%     % Epsilon transitions back to q0
%     (q1) edge[bend left=30] node[below,midway] {$\varepsilon, \lambda\varepsilon$} (q0)
%     (q2) edge[bend right=20] node[right,midway] {$\varepsilon, \lambda\varepsilon$} (q0);
% \end{tikzpicture}
% \caption{SFT representation of the HTML encoding sanitizer function.}
% \label{fig:html-sanitizer-sft1}
% \end{figure}

We first evaluate the performance of our SFT product operation by computing the product of the HTML encoding sanitizer SFT shown in Figure \ref{fig:html-sanitizer-sft} (see Section \ref{sec:motivation} for motivation) with SFAs constructed from randomly generated Unicode sequences. This product represents the sanitization result produced by the SFT. The experimental evaluation was conducted on a laptop equipped with an Apple M4 processor and 24 GB of memory, with a one-minute time limit imposed for each test.

@@

We also compare our approach with two state-of-the-art string solvers: Ostrich \cite{pacmpl/ChenFHHHKLRW22} and CVC5 \cite{cvc5}.
The experimental results demonstrate that string replacement operations, particularly replace-all operations, pose significant challenges for existing state-of-the-art solvers.
The "unknown" label in the table indicates instances where CVC5 cannot determine satisfiability. 
Notably, PCP and RNA benchmarks cannot be solved by most state-of-the-art string
solvers, including CVC5 \cite{cvc5} and Z3Str \cite{Z3-str3}, yet CertiStrR solves
them efficiently. Among existing solvers, only OSTRICH \cite{pacmpl/ChenFHHHKLRW22} can handle these benchmarks.
CertiStrR outperforms both Ostrich and CVC5 across most benchmark subgroups, with the only exception of the "Replace Str" category where CVC5 achieves better performance (though it experiences one timeout). CertiStrR demonstrates particularly strong performance in the "Replace All" category.

@@
